---
title: "Analyzing the Gapminder Dataset"
author: "By: Mwangi George"
date: "Last edited: `r format(Sys.Date(), '%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    theme: cerulean
---

## Introduction
In this paper, I am going to run SQL queries with the `gapminder` dataset. The goal is to showcase my understanding of the following SQL clauses:
 
 * SELECT
 * WHERE
 * ORDER BY 
 * GROUP BY
 * ALIASING

### Loading Important Packages and Creating database connection
```{r}
# loading packages
pacman::p_load(tidyverse, gapminder, RSQLite, DBI)

# Create database connection
lite_connect <- dbConnect(SQLite(), "myProjectDatabase.sqlite")

```

### Loading Gapminder dataset and Uploading it to database
```{r}
# load dataset into R environment
gapminder <- gapminder

# Write the data to the database as a table
dbWriteTable(conn = lite_connect, 
             name = "gapminder", 
             value = gapminder,
             overwrite = T)

# Check whether the dataset now exists in the database
dbExistsTable(conn = lite_connect,
              name = "gapminder")

# select the last 5 rows from the gapminder table
dbGetQuery(conn = lite_connect,
           statement = "SELECT * FROM gapminder LIMIT 5")
```

## SQL Statements

### Write a query to extract only the observations from the year 1957
```{sql connection = lite_connect}
SELECT * 
FROM gapminder
WHERE year = 1957
```

### Filter the gapminder data to retrieve only the observation from China in the year 2002
```{sql connection = lite_connect}
SELECT * 
FROM gapminder 
WHERE country = "China" 
AND year = 2002
```

### Sort the gapminder dataset in ascending order of life expectancy (lifeExp)
```{sql connection = lite_connect}
SELECT * 
FROM gapminder 
ORDER BY lifeExp
```

### Write a query to retrieve all observations from the gapminder table in descending order of life expectancy (lifeExp)
```{sql connection = lite_connect}
SELECT * 
FROM gapminder 
ORDER BY lifeExp 
DESC
```

### Write a query to extract observations from just the year 1957 sorted in descending order of population (pop)
```{sql connection = lite_connect}
SELECT * 
FROM gapminder 
WHERE year = 1957 
ORDER BY pop 
DESC
```


### Write a query that returns the distinct number of countries in the Americas continent. Name the results `no_of_countries_in_America`
```{sql connection = lite_connect}
SELECT  COUNT(DISTINCT country)
        AS no_of_countries_in_America
FROM gapminder 
WHERE continent = "Americas"
```

### Write a query that returns the all records for the United States for 1997, 2002, and 2007
```{sql connection = lite_connect}
SELECT * 
FROM gapminder 
WHERE  continent = "Americas"
AND country = "United States" 
AND year IN (1997, 2002, 2007)
```

### Write a query that returns the average life expectancy in the United States as `avg_lifeExp_US` for 2007
```{sql connection = lite_connect}
SELECT AVG(lifeExp) 
       AS avg_lifeExp_US 
FROM gapminder 
WHERE continent = "Americas"
AND country = "United States"
AND year = 2007
```

### Write a query to calculate the average life expectancy per continent as `avg_lifeExp` in 2007
```{sql connection = lite_connect}
SELECT  continent, 
        AVG(lifeExp) AS avg_lifeExp
FROM gapminder
WHERE year = 2007
GROUP BY continent
```

### Write a query to calculate the total population per continent as `total_pop` in 2007. Sort the results in decreasing order of total population
```{sql connection = lite_connect}
SELECT  continent,
        SUM(pop) AS total_pop
FROM gapminder
WHERE year = 2007
GROUP BY continent
ORDER BY total_pop DESC
```

### Write a query that returns the country, year, population in millions of people, rounded to two decimal places and life expectancy rounded to one decimal place.
```{sql connection = lite_connect}
SELECT  country,
        year, 
        ROUND(pop/1000000.0, 2) AS pop_in_millions,
        ROUND(lifeExp, 1) AS lifeExp
FROM gapminder
```

### Write a query that returns the country, year, life expectancy and population in thousands of people for any field with a life expectancy greater than 70
```{sql connection = lite_connect}
SELECT  country, 
        year, 
        lifeExp,
        pop/1000.0
FROM gapminder
WHERE lifeEXP > 70
```
There are 493 records where life expectancy is greater than 70.

### How many records are there if we change lifeExp to greater than 75?
```{sql connection = lite_connect}
SELECT  country, 
        year, 
        lifeExp,
        pop/1000.0
FROM gapminder
WHERE lifeEXP > 75
```
There are 173 records where life expectancy is greater than 75.

### Write a query that returns the country, year, life expectancy and population in thousands of people for any field with a life expectancy greater than 70 and before 1990. How many records are there?
```{sql connection = lite_connect}
SELECT  country,
        year, 
        lifeExp,
        pop/1000.0 AS pop_in_millions
FROM gapminder
WHERE lifeExp > 70 
AND year > 1990
```
There are 277 records.

### How many records are there if we just look at the year 1952?
```{sql connection = lite_connect}
SELECT  country,
        year, 
        lifeExp,
        pop/1000.0 AS pop_in_millions
FROM gapminder
WHERE lifeExp > 70 
AND year = 1952

```
There are 5 records if we just look at the year 1952.

### How many records are there if we just look at the year 2007?
```{sql connection = lite_connect}
SELECT  country,
        year, 
        lifeExp,
        pop/1000.0 AS pop_in_millions
FROM gapminder
WHERE lifeExp > 70 
AND year = 2007

```
There are 83 records if we just look at the year 2007.

```{r}
# disconnect database
dbDisconnect(lite_connect)
```

